Do researchers preferentially collaborate with colleagues of the same gender?

### 2018-06-03 manuscript/Manuscript_Claire.Rmd
Merge in Claire's updates

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious
  biases, and often feel excluded from formal and informal opportunities for research
  collaboration. In addition to ensuring fairness and helping to redress gender imbalance
  in the academic workforce, increasing women's access to collaboration could help
  scientific progress by drawing on more of the available human capital. Here, we
  test whether researchers preferentially collaborate with same-gendered colleagues,
  using more stringent methods and a larger dataset than in past work. Our results
  reaffirm that researchers co-publish with colleagues of the same gender, and show
  that this 'gender homophily' is slightly stronger today than it was 10 years ago.
  Contrary to our expectations, we found no evidence that homophily is driven mostly
  by senior academics, and no evidence that homophily is strongest in fields where
  women are in the minority. Interestingly, homophily was negatively correlated with
  journal impact factor, as predicted if mixed-gender teams produce better research.
  We discuss potential causes of gender homophily in science. \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining,
  Women in STEM. 
---

@@

Women are severely underrepresented in many branches of science, technology, engineering, mathematics, and medicine (STEMM), and face additional challenges and inequities relative to men [e.g. @Shaw_2012; @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions [@Wutte_2007; @Reuben_2014] with lower salaries [@Trower_2002; @Umbach_2007], receive less grant money [@Hosek_2005; @OER_2008], are promoted more slowly [@Zuckerman_1987; @Rosenfeld_1991; @Long_1993; @Hopkins_2013], and are allocated fewer resources [@ODorchai_2009] and less research funding [@Feldt_1986; @Stack_2004; @Lariviere_2011]. Experimental evidence suggests that bias against women plays a major role in generating these differences [@Moss_2012; @Knobloch_2013].

@@

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008]. Additionally, empirical studies imply that mixed-gender or otherwise diverse teams produce better results on collaborative tasks than less diverse teams [@reagans2001; @Hong_2004; @Bear_2011; @Campbell_2013]. For reasons such as these, several studies have examined the author lists of published research articles in order to test for gender differences in collaboration frequency or pattern. To our knowledge, all of these studies imply that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing is often termed 'gender homophily'.

@@

In the present study, we test whether life sciences researchers tend to co-publish with same-gendered colleagues, while controlling for the Wahlund effect as strictly as possible. We use a recently-published dataset describing the gender of 35.5m authors from 9.15m articles indexed on PubMed [@Holman_2018]. Holman et al. [-@Holman_2018] reported large differences in the gender ratio of authors across research disciplines, journals, countries, and across the years 2002-2016. We therefore tested for gender homophily after restricting each analysis to particular journals (i.e. research specialties), time periods, and countries.

# Methods
## The dataset
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every article indexed on PubMed and attempting to infer each author's gender from their given name. Each journal was assigned to one of 107 scientific disciplines, using PubMed's journal categorisations in the interests of objectivity. Because the present study focuses on co-authorship, all single-author papers were discarded. We also discarded all papers for which we could not determine the gender of every author with ${\ge}95\%$ certainty, in order to simplify the statistical analysis. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published either 0-1 or 10-11 years before the PubMed data were collected (i.e. 20^th^ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers.

The final sample sizes were `r sample.size$n2015[3]` papers and `r sample.size$n2015[4]` authorships from August 2015 - August 2016, and `r sample.size$n2005[3]` papers and `r sample.size$n2005[4]` authorships from August 2005 - August 2006. Respectively, these papers came from `r sample.size$n2015[2]` and `r sample.size$n2005[2]` journals, which were grouped into `r sample.size$n2015[1]` and `r sample.size$n2005[1]` research disciplines. There was a median of `r sample.size$n2015[5]` (`r sample.size$n2005[5]`) papers per journal, `r sample.size$n2015[6]` (`r sample.size$n2005[6]`) authors per journal, and `r sample.size$n2015[7]` (`r sample.size$n2005[7]`) authors per paper (the first number is for 2015-6, and the bracketed number for 2005-6).

## Calculating $\alpha$, the coefficient of homophily
Following Bergstrom et al. [-@bergstrom_2016], we defined the coefficient of homophily as $\alpha = p - q$, where $p$ is the probability that a randomly-chosen co-author of a _male_ author is a man and $q$ is the probability that a randomly-chosen co-author of a _female_ author is a man. Therefore, $\alpha > 0$ suggests that same-gender authors publish together more often than expected under the null hypothesis that authors assort randomly with respect to gender (homophily), $\alpha < 0$ suggests that opposite-gender authors publish together more often than expected (heterophily), and $\alpha = 0$ suggests random assortment with respect to gender. Like the Wahlund effect, $\alpha$ is borrowed from population genetics; for a set of 2-author papers, it is equivalent to Wright's coefficient of inbreeding [@wright_1949]. Mathematical work illustrates that $\alpha$ is closely related to network-based alternative methods for quantifying homophily [@wang_2016].

@@

As expected, $\alpha'$ was usually almost identical to $\alpha$ (Figure S1), but $\alpha$ was downwardly biased relative to $\alpha'$ for small datasets (Figure S2). Additionally, the correlation between $\alpha'$ and sample size was negligible ($R^2 < 0.01$), suggesting that our calculation of $\alpha'$ effectively removed the dependence of $\alpha$ on sample size. We therefore used $\alpha'$ in all analyses.

@@

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha' > 0$, and for many of these the FDR-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6. The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender.

![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after adjusting p-values using Benjamini-Hochberg false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/figure2.pdf)

$\alpha'$ was significantly higher in the literature sample from 2015-16 relative to 2005-6, though the difference in means was small (Figure S3; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.091{\pm}0.04$, $t_{953}$ = 2.42, p = 0.016).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor 'Authorship position' in a linear mixed model: Cohen's $d$ = $0.065{\pm}0.02$, $t_{2024}$ = 4.28, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors' genders than the gender of the last author, i.e. the opposite of what is predicted if senior scientists are causally responsible for homophily.

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, Table S3). However, there was no evidence for consistent differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained around 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). Thus, we cannot reject the null hypothesis that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

## Analysis accounting for differences in author gender ratio between countries
When we restricted the analysis by country, we observed statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant evidence of heterophily (Figures S5-S6). Additionally, the values of $\alpha'$ calculated for each journal-country combination were mostly very similar to the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries); on average, $\alpha'$ was only inflated in the main (non-country specific) analysis by 0.002 (Figure S7). These results suggest that our findings of widespread homophily in the main analysis were not driven solely by a Wahlund effect resulting from gender differences between countries.

We also found that gender homophily was marginally stronger in 2015-2016 relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators over time, there are alternative (and perhaps more likely) explanations. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Shaw_2012; @Long_2015; @McKenzie_2017; @Bendels_2018]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha$ may increase.

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@reagans2001; @Hong_2004; @Bear_2011; @Campbell_2013].

Homophily might also have more benign causes. Collaboration is often most enjoyable and productive when working with like-minded people, who might be same-gendered more often than not. We also suppose that some people consciously choose to preferentially collaborate with women in order to help close the gender gap in the workforce; this would create homophily if women do this more than men. In support of this interpretation, women appear more likely than men to promote the work of female collagues by inviting them to give talks [@nittrouer_2018; @debarre_2018]. Related to this, Ghiasi et al. [-@Ghiasi_2015] concluded that women in engineering are "compliant [in reproducing] male-dominated scientific structures" because they do not collaborate often enough with other women; their data suggest that coauthorships between two women are about 30% more frequent than expected under the null (see their Figure 7). In contrast, we propose that men as well as women should ensure that they are not inadvertently overlooking or excluding female colleagues, particularly since men are disproportionately represented among senior researchers [@Shaw_2012].

----------

Why does gender homophily exist in science? Collaboration often leads to co-authorship, publications, citations and additional professional recognition, therefore first authors should always be considering collaborators that are most likely to enhance their productivity and in the long term also their academic success, rather than pick co-authors according to their gender. First of all, it is likely that decades ago, the probability to meet a female colleague working on the same area of expertise was slimmer than now. Even though the number of women in STEMM is increasing, i.e. it is not likely to be an obstacle to collaboration anymore; we cannot forgo that female representation in science is still not equal to that of males. Second, the most basic source of homophily is space, scientists are more likely to collaborate with colleagues closer to them in geographic location (@McPherson_2011). In the case of women, personal considerations (e.g. family commitments) represent the principal restrictions on their international mobility. As a result women are less likely to originate collaborations with foreign colleagues [@Lemoine_1992; @Lewison_2001; @Webster_2001; @Lariviere_2011]. Lastly, men and women think differently on average (REF) and are more likely to initiate collaborations based on similarity of thoughts to progress their scientific output (REF).

Academic publishing is one of the most (if not the most) important tool for a researcher to display their work and is frequently used as an indicator of success for being promoted and hired as a faculty member. Our finding that researchers are more likely to publish with same-gendered co-authors implies that co-authorship listing is not a random process, but instead creates additional gender imbalance. Homophily could make it harder for women to find collaborators in disciplines where women are in the minority [i.e. almost all STEMM fields; @Holman_2018]. Because women make up for as little as 30% of the STEMM scientist community [@Holman_2018], they are less likely to be picked as coauthors, and, as a result, gender homophily creates further gender imbalance and disadvantages for women. Men tend to publish more in high impact journals as first author [@Holman_2018], and generate more citations than papers lead by first author women [@Long_1992; @Lariviere_2013]. As a consequence, gender homophily is also likely to contribute to lower publication rates for women [@McDowell_1992].

Although the proportion of women in science has significantly increased over the last decades, women still tend to leave their academic careers behind considerably more often than their male counterparts [REF]. Lack of scientific recognition [@Lincoln_2012], lack of research funding [@Bozeman_2004], lack of invitation to speak at scientific events [@Isbell_2012; @Klein_2017; @debarre_2018], discrimination during hiring [@MossRacusin_2017; @Reuben_2014] and discrimination when writing a recommendation letter [@Trix_2003; @Schmader_2007], are only some of the small disadvantages adding up to drive women out of science. Whether their decision to leave academia is deliberate or not, it will negatively impact science overall, as mentoring, networking, and supporting younger women scientists are especially important for women in academia.

Altogether, our findings suggest that researchers preferably collaborate with same-gendered colleagues in STEMM research. Despites using a more stringent analysis method (e.g. Wahlund effects) as well as the largest dataset to date, our results were in accord with longstanding and recent research, while adding new details, reasoning and insight for the current situation. Gender homophily in science not only reinforces the less favorable status of women in science, but also creates long term problems as described above. It is paramount to explore the causes and implement measures to readdress the gender disparity. We advocate for the introduction and implementation of gender policies in order to support opposite sex collaboration, to give men and women a chance to collaborate across gender, fields, and to promote scientific and innovative excellence.

# Acknowledgements
We are grateful to XXXX. CM was supported by the Academy of Finland (284666 to the Centre of Excellence in Biological Interactions).

@@

>>>>

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious
  biases, and often feel excluded from formal and informal opportunities for research
  collaboration. In addition to ensuring fairness and helping to redress gender imbalance
  in the academic workforce, increasing women's access to collaboration could help
  scientific progress by drawing on more of the available human capital. Here, we
  test whether researchers preferentially collaborate with same-gendered colleagues,
  using more stringent methods and a larger dataset than in past work. Our results
  reaffirm that researchers co-publish with colleagues of the same gender, and show
  that this 'gender homophily' is slightly stronger today than it was 10 years ago.
  Contrary to our expectations, we found no evidence that homophily is driven mostly
  by senior academics, and no evidence that homophily is strongest in fields where
  women are in the minority. Interestingly, journals with a high impact factor for their discipline tended to have comparatively low homophily, as predicted if mixed-gender teams produce better research.
  We discuss potential causes of gender homophily in science. \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining,
  Women in STEM. 
---

@@

Women are severely underrepresented in many branches of science, technology, engineering, mathematics, and medicine (STEMM), and face additional challenges and inequities relative to men [e.g. @Shaw_2012; @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions [@Wutte_2007; @Reuben_2014] with lower salaries [@Trower_2002; @Umbach_2007], receive less grant money [@Hosek_2005; @pohlhaus_2011], are promoted more slowly [@Zuckerman_1987; @Rosenfeld_1991; @Long_1993; @Hopkins_2013], and are allocated fewer resources [@ODorchai_2009] and less research funding [@Feldt_1986; @Stack_2004; @Lariviere_2011]. Experimental evidence suggests that bias against women plays a major role in generating these differences [@Moss_2012; @Knobloch_2013].

@@

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008]. Additionally, empirical studies imply that mixed-gender or otherwise diverse teams produce better results on collaborative tasks than less diverse teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. For reasons such as these, several studies have examined the author lists of published research articles in order to test for gender differences in collaboration frequency or pattern. To our knowledge, all of these studies imply that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing is often termed 'gender homophily'.

@@

In the present study, we test whether life sciences researchers tend to co-publish with same-gendered colleagues, while controlling for the Wahlund effect as strictly as possible. We use a recently-published dataset describing the gender of 35.5m authors from 9.15m articles indexed on PubMed [@Holman_2018]. Holman et al. [-@Holman_2018] reported large differences in the gender ratio of authors across research disciplines, journals, countries, and across the years 2002-2016. We therefore tested for gender homophily while restricting our analysis to particular journals (i.e. research specialties), time periods, and countries. We quantified gender assortment using a metric called $\alpha'$ [@bergstrom_2016], which is positive when same-gender authors publish together more often than expected (gender homohily), negative when opposite-gender authors publish together more often than expected (heterophily), and equal to zero when authors assort randomly with respect to gender (see Methods).

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha' > 0$, and for many of these the FDR-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender.

![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of gender homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/figure2.pdf)

$\alpha'$ was significantly higher in the literature sample from 2015-16 relative to 2005-6, though the difference in means was small (S1 Fig; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.091{\pm}0.04$, $t_{953}$ = 2.42, p = 0.016).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (S2 Fig; Effect of the fixed factor 'Authorship position' in a linear mixed model: Cohen's $d$ = $0.065{\pm}0.02$, $t_{2024}$ = 4.28, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors' genders than the gender of the last author, i.e. the opposite of what is predicted if senior scientists are causally responsible for homophily.

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, S3 Table). However, there was no evidence for consistent differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained around 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). Thus, we cannot reject the null hypothesis that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

## Analysis accounting for differences in author gender ratio between countries
When we restricted the analysis by country, we observed statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant evidence of heterophily (Figures S3-S4). Additionally, the values of $\alpha'$ calculated for each journal-country combination were mostly very similar to the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries); on average, $\alpha'$ was only inflated in the main (non-country specific) analysis by 0.002 (S5 Fig). These results suggest that our findings of widespread homophily in the main analysis were not driven solely by a Wahlund effect resulting from gender differences between countries.

We also found that gender homophily was marginally stronger in 2015-2016 relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators over time, there are alternative (and perhaps more likely) explanations. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Long_2015; @Bendels_2018; @McKenzie_2017]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha$ may increase.

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013].

Homophily might also have more benign causes. Collaboration is often most enjoyable and productive when working with like-minded people, who might be same-gendered more often than not. We also suppose that some people consciously choose to preferentially collaborate with women in order to help close the gender gap in the workforce; this would create homophily if women do this more than men. In support of this interpretation, women appear more likely than men to promote the work of female collagues by inviting them to give talks [@nittrouer_2018; @debarre_2018]. Related to this, Ghiasi et al. [-@Ghiasi_2015] concluded that women in engineering are "compliant [in reproducing] male-dominated scientific structures" because they do not collaborate often enough with other women; their data suggest that coauthorships between two women are about 30% more frequent than expected under random assortment (see their Figure 7). In contrast, we propose that men as well as women should ensure that they are not inadvertently overlooking or excluding female colleagues, particularly since men are disproportionately represented among senior researchers [@Shaw_2012].

# Methods
## The dataset
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every article indexed on PubMed and attempting to infer each author's gender from their given name. Each journal was assigned to one of 107 scientific disciplines, using PubMed's journal categorisations in the interests of objectivity. Because the present study focuses on co-authorship, all single-author papers were discarded. We also discarded all papers for which we could not determine the gender of every author with ${\ge}95\%$ certainty, in order to simplify the statistical analysis. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published either 0-1 or 10-11 years before the PubMed data were collected (i.e. 20^th^ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers. Detailed sample size information is given in S1 Table.

## Calculating $\alpha$, the coefficient of homophily
Following Bergstrom et al. [-@bergstrom_2016], we defined the coefficient of homophily as $\alpha = p - q$, where $p$ is the probability that a randomly-chosen co-author of a _male_ author is a man and $q$ is the probability that a randomly-chosen co-author of a _female_ author is a man. Like the Wahlund effect, $\alpha$ is borrowed from population genetics; for a set of 2-author papers, it is equivalent to Wright's coefficient of inbreeding [@wright_1949]. Mathematical work illustrates that $\alpha$ is closely related to alternative network-based methods for quantifying homophily [@wang_2016].

@@

As expected, $\alpha'$ was usually almost identical to $\alpha$ (S6 Fig), but $\alpha$ was downwardly biased relative to $\alpha'$ for small datasets (S7 Fig). Additionally, the correlation between $\alpha'$ and sample size was negligible ($R^2 < 0.01$), suggesting that our calculation of $\alpha'$ effectively removed the dependence of $\alpha$ on sample size. We therefore used $\alpha'$ in all analyses.

@@

# Acknowledgements
CM was supported by the Academy of Finland (284666 to the Centre of Excellence in Biological Interactions).

@@

\newpage
# Supporting information
S1 Fig.  There is a very strong correlation between the values of $\alpha$ and $\alpha'$ calculated for each journal, though in a handful of cases the difference is considerable. The deviation between $\alpha$ and $\alpha'$ is greatest for journals for which there is a small sample size (see S2 Fig).

S2 Fig. For journals for which we recovered a small number of papers (<100), the unadjusted metric $\alpha$ was downwardly biased. This fits our expectations: because authors cannot be their own co-authors, small datasets will tend to produce negative estimates of $\alpha$ even if authors assort randomly with respect to gender (see main text). This suggests that $\alpha'$ is a more useful measure of homophily and heterophily, especially for small samples.

S3 Fig. Histogram showing the distribution of differences in $\alpha'$ between the 2015-16 and 2005-6 samples, where positive numbers indicate an increase in $\alpha'$ with time. The mean is slightly positive (namely 0.004), indicating a mild increase in average $\alpha'$ with time.

S4 Fig. Histogram showing the difference between $\alpha'$ calculated for first and last authors. Positive values mean that $\alpha'$ was higher when calculated for first authors, and negative values mean $\alpha'$ was higher when calculated for last authors. The mean is very slightly higher than zero, indicating that $\alpha'$ tends to be higher for first authors.

S5 Fig. Histogram of $\alpha'$ for 325 unique combinations of journal and country, using data from August 2015 - August 2016. The white areas denote combinations for which $\alpha'$ differs significantly from zero (p < 0.05, following false discovery rate correction).

S6 Fig. Plot showing the 68 combinations of journal and author country of affiliation for which $\alpha'$ is significantly higher than expected.

S7 Fig. Histogram showing the estimated degree to which $\alpha'$ is inflated by inter-country differences in author gender ratio, across the 285 journals for which we had adequate data after restricting the analysis by country. The average inflation in $\alpha'$ is negligible, suggesting that Wahlund effects resulting from inter-country differences have a neglible effect on our estimates of gender homophily.

S1 Table. The $\alpha'$ values calculated for each journal, in the 2005 and 2015 samples, for each type of author (i.e. all authors, first authors, and last authors). The tables gives the impact factor of each journal, the sample size, $\alpha$ and $\alpha'$ and their 95% CIs, and the p-value from a 2-tailed test evaluating the null hypothesis that $\alpha$ is zero (both raw and FDR-corrected p-values are shown).

S2 Table. Number of journals showing statistically significant homophily or heterophily, in two one-year periods. The significance threshold was p < 0.05, and p-values were adjused using Benjamini-Hochberg false discovery rate correction. Note that the power of our test is lower for the 2005-2006 data because fewer papers were recovered per journal: thus, it is not meaningful to compare the % significant journals (i.e. 11% vs 24%) between the two time periods.

S3 Table. The table shows similar information to S3 Fig, namely the distribution of the $\alpha'$ values across journals, split by the research discipline. The gender ratio column shows the percentage of women authors in the sample used to calculate $\alpha'$, across all authorship positions. In the last two columns, the numbers outside parentheses give the number of journals that deviate statistically significantly from zero, while the numbers inside parentheses give the number that remain significant after false discovery rate correction.
